---
title: "SWIB24: Redlining"
date: "Redlining, Health and the Environment"

fontsize: 11pt
geometry: margin=1in

output:
  pdf_document:
    includes:
      in_header: header.tex
    fig_width: 5
    fig_height: 3.5
---

\textbf{Reading on graphical summaries}      

- *OI Biostat*: Section 1.6 and earlier material on numerical and graphical summaries.

\vspace{1cm}

\textbf{Overview} 

We will use data compiled in this tutorial available online to assess the effects of historical Redlining with modern day outcomes. \url{https://github.com/Jayanth-Mani/Redlining_Data_Tutorial/tree/master}. 

### Redlining 
Redlining refers to the the 1930s practice by the Home Owners’ Loan Corporation (HOLC), which drew lines on maps  to grade loan security on an A-D scale. These maps were used to limit 
access to mortgages in predominantly African American and immigrant neighborhoods. Previous research has shown that formerly redlined neighborhoods have continuing modern-day impacts 
on health, environment such as worse air pollution and lower economic activity. 

Please read this New York Times article \url{https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html}

## Data
The compiled data provided merges information from several data sources listed below: 

1. [The American Community Census](https://www.census.gov/programs-surveys/acs)
2. [University of Richmond HOLC dataset (2010 census)](https://github.com/americanpanorama/Census_HOLC_Research/tree/main/2010_Census_Tracts)
3.    [Diversity Data Kids HOLC grading mapped to the 2010 census](https://data.diversitydatakids.org/dataset/holc_census_tracts-home-owner-loan-corporation--holc--neighborhood-grades-for-us-census-tracts?_ga=2.95776788.927340397.1689780461-135717849.1688844078)
4.    [EJScreen Census Tract level data](https://gaftp.epa.gov/EJScreen/2022/EJSCREEN_2022_StatePct_with_AS_CNMI_GU_VI_Tracts.csv.zip)
5. [Center for Air, Climate and Energy Solutions (CACES) Air Pollution data](https://www.caces.us/data)
6.    [USDA Food Access (2019 Food Atlas)](https://www.ers.usda.gov/data-products/food-access-research-atlas/download-the-data/#Current%20Version)
7.    [Diversity Data Kids Child Opportunity Index](https://data.diversitydatakids.org/dataset/coi20-child-opportunity-index-2-0-database/resource/080cfe52-90aa-4925-beaa-90efb04ab7fb)
8. [Opportunity Atlas](https://www.opportunityatlas.org/)
9. [Open Park Area from the National Neighborhood Data Archive (NaNDA)](https://www.openicpsr.org/openicpsr/project/117921/version/V1/view)


## Some code to help you get started 

1. The file we are reading is a geojson file which includes spatial information corresponding to census tracts in the US. We will use specific R packages to read this type of data and define some important variables that can help you in your analysis. 

```{r}
setwd("~/Documents/School Stuff/Biostatistics/Redlining Project")
library(sf)
library(spdep)
library(class)
library(dplyr)
library(ggplot2)
library(ggfortify)



mydat <- st_read("final_df.geojson")
dim(mydat)

# mydat <- na.omit(mydat)
```


```{r}
## Removing census tracts with invalid spatial coordinates
table(st_is_valid(mydat))

analysis.dat <- mydat[st_is_valid(mydat),]

datMA <- analysis.dat[analysis.dat$STATE_NAME == "Massachusetts", ]


## Defining key variables
## HOLC grade: A=best, D=worst

grade <- analysis.dat$class1 
table(grade)
sort(table(analysis.dat$STATE_NAME))

## Some key demographic variables at the census tract level 
## These are demographics are from 2015 ACS survey, which releases 
## aggregated data collected over the previous 5 year period from random 
## samples of households across the US. 

## total population
population <- analysis.dat$pop_num
hist(population)

## median age
age <- analysis.dat$median_age
hist(age)


## % minorities
minoritypct  <- analysis.dat$minority_pct.x
hist(minoritypct)

## % unemployed
unins <- analysis.dat$uninsured_pct
hist(unins)

## median family income
income <- analysis.dat$MedianFamilyIncome
income <- as.numeric(income)




```

## Data on Food Deserts 
The LILA variables (eg. LILATracts_1And10) are indicators for food deserts using a variety of definitions. For example, LILATracts_1And10 is a flag for food desert when considering low accessibilty at 1 and 10 miles. 

Labels for all columns starting with "la.." can be found here: #https://data-dictionary.regenstrief.org/iadc/catalog/s#howDataset/IADC/4

```{r}
## LILA10

lila10 <- analysis.dat$LILATracts_1And10
```

## Health outcomes 
Census tract level summaries of specific health outcomes shown in the code below was obtained from \url{https://dsl.richmond.edu/panorama/redlining/data}

```{r}

## Life expectancy
le <- analysis.dat$life_exp

## Prevalence of high blood pressure 
bp <- analysis.dat$highbp_pct

## Prevalence of cancer 
cancer <- analysis.dat$cancer_pct

## Prevalence of asthma
asthma <- analysis.dat$asthma_pct

## Prevalence of coronary heart diseae 
chd <- analysis.dat$chd_pct

## Prevalence of COPD 
copd <- analysis.dat$copd_pct

## and others in columns 20:24
```

## Air pollution estimates from 2010 

Air pollution estimates including carbon monoxide, nitrogen dioxide, ozone, PM10 and sulphur dioxide are from the Center for Air, Climate and Energy Solutions (CACES) Air Pollution (\url{https://www.caces.us/data}) 
```{r}
# carbon monoxide
co <- analysis.dat$co

# nitrogen dioxide
no2 <- analysis.dat$no2

# ozone 
o3 <- analysis.dat$o3

# PM10 fine particular matter 
p10 <- analysis.dat$pm10

# And other pollutants
```

## Green space 

Data on open park lands is from the Open Park Area from the National Neighborhood Data Archive (https://www.openicpsr.org/openicpsr/project/117921/version/V1/view)

## Variable names 

There are 427 variables (columns) in this dataset. A guide to the variables and their source is in the excel spreadsheet Column_Labels.xlsx. 

## US States represented

Not all states are equally represented. If modern day census tracts were not mapped to HOLC areas, they won't be included in these data. If you would like to focus your analysis within a state, this will be helpful. You can see that California, New York and Illinois have data on more than 1000 census tracts

```{r}
sort(table(analysis.dat$STATE_NAME))

```

## Code for a spatial plot
You can make cool spatial plots using ggplot2 in R. See map of Sussex County, MA (Boston and surrounding cities/towns). 

```{r}
library(ggplot2)
#install.packages("wesanderson")

## subset data for MA
datMA <- analysis.dat[analysis.dat$STATE_NAME == "Massachusetts", ]

## fill in missing values for income by setting it to equal median income across MA
datMA$MedianFamilyIncome <- as.numeric(datMA$MedianFamilyIncome)
datMA[is.na(datMA$MedianFamilyIncome)] <- median(datMA$MedianFamilyIncome, na.rm=T)

## ggplot2 for spatial plot (the wesanderson pkg gives you the cool colors!)
ggplot(data=datMA[datMA$County == 25,]) + geom_sf(aes(fill=MedianFamilyIncome)) + scale_fill_gradientn(colors=wesanderson::wes_palette("Zissou1", 100, type = "continuous"))

# Assuming 'datMA' already contains your data

# Create a mapping for the classes
class_mapping <- c("A" = 1, "B" = 2, "C" = 3, "D" = 4)

# Add a new column 'class_numeric' based on 'class1'
datMA$class_numeric <- class_mapping[datMA$class1]

# Show the updated dataset
head(datMA)

## compare to plot colored by HOLC grade 

```

```{r}
library(stats)
library(randomForest)
library(wesanderson)
library(ggplot2)
library(cluster)

ntrees <- 500

set.seed(123)  
# Filter out rows where asthma_pct is zero
datMA_filtered <- subset(datMA, asthma_pct != 0)

library(tidyverse)

health_data = datMA_filtered %>% 
  select(life_exp, highbp_pct, cancer_pct, asthma_pct, chd_pct, copd_pct, diabetes_pct, kidney_pct, mental_health_pct, obesity_pct, stroke_pct)

econ_data = datMA_filtered %>% 
  select(z_oi, MedianFamilyIncome, sovi, poverty_pct, uninsured_pct, UNEMPPCT, LESSHSPCT)

health_data = as.matrix(health_data[,-12])
# Train the random forest model with the filtered dataset
health_model <- kmeans(
  health_data[,-12],
  centers = 4
)

econ_model <- kmeans(
  health_data[,-12],
  centers = 4
)

print(health_model)
datMA_filtered$health_cluster <- as.factor(health_model$cluster)

datMA_filtered$econ_cluster <- as.factor(econ_model$cluster)

health_confusion_matrix <- table(datMA_filtered$health_cluster, datMA_filtered$class_numeric)

# Print the confusion matrix
print(health_confusion_matrix)


# ggplot2 for spatial plot for health data
ggplot(data = datMA_filtered[datMA_filtered$County == 25,]) +
  geom_sf(aes(fill = health_cluster)) +
  scale_fill_manual(values = wes_palette("Zissou1", 4, type = "continuous")) +
  labs(title = "K-means Clustering of Health Data by County",
       fill = "Cluster") +
  theme_minimal()
# ggplot2 for spatial plot for econ data
ggplot(data = datMA_filtered[datMA_filtered$County == 25,]) +
  geom_sf(aes(fill = econ_cluster)) +
  scale_fill_manual(values = wes_palette("Zissou1", 4, type = "continuous")) +
  labs(title = "K-means Clustering of Socioeconomic Data by County",
       fill = "Cluster") +
  theme_minimal()


```

```
econ_data = datMA_filtered %>% 
  select(z_oi, MedianFamilyIncome, sovi, poverty_pct, uninsured_pct, UNEMPPCT, LESSHSPCT{r}


```


